---
title: "week4Assignment"
output: html_document
---


This database is known as the National Emissions Inventory (NEI). You can read more information about the NEI at the EPA National Emissions Inventory web site.

For each year and for each type of PM source, the NEI records how many tons of PM2.5 were emitted from that source over the course of the entire year. The data that you will use for this assignment are for 1999, 2002, 2005, and 2008.

Data for assignment:
The zip file contains two files:

PM2.5 Emissions Data (summarySCC_PM25.rds): This file contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment

The overall goal of this assignment is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999-2008. You may use any R package you want to support your analysis.

For each question/task you will need to make a single plot. Unless specified, you can use any plotting system in R to make your plot.

```{r Read Data}

temp <- tempfile()
temp2 <-tempfile()

download.file("https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip", temp)
unzip(temp,list=TRUE)
unzip(zipfile=temp, exdir=temp2)
list.files(temp2)

## This first line will likely take a few seconds. Be patient!
NEI <- readRDS(file.path(temp2,'summarySCC_PM25.rds',fsep='\\'))
SCC <- readRDS(file.path(temp2,"Source_Classification_Code.rds",fsep='\\'))

```

### Question 1

Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

```{r Question 1}

# Emissions: Amount of PM2.5 emitted, in tons

# find total of pm2.5 emission from all sources for each of the years 1999, 2002, 2005, 2008

a <- split(NEI, NEI$year)
emission_summary <- sapply(a, function(grp) sum(grp$Emissions, na.rm=TRUE))

#create a plot, xaxt to hide the x axis label, ann to hide axis labels
plot(emission_summary, xaxt="n", ann=FALSE, col = names(emission_summary))
title(main = "Annual emission of pm2.5", xlab="year", ylab='pm2.5 amount')
axis(1, at=1:4, labels=names(emission_summary))


png("Plot1.png")

barplot(emission_summary, ann=FALSE, col = names(emission_summary), names.arg=names(emission_summary))
title(main = "Annual emission of pm2.5", xlab="year", ylab='pm2.5 amount')

dev.off()

```

### Question 2
Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.

```{r Question 2}

# subset the data
baltimore <- subset(NEI,fips=="24510")

emission_summary <- sapply(split(baltimore,baltimore$year), function(grp) sum(grp$Emissions, na.rm=TRUE))

png("Plot2.png")
    plot(emission_summary,pch = 21, cex=4,bg=names(emission_summary), xaxt="n", ann=FALSE)
    title(main = "Annual emission of pm2.5 in Baltimore City, Maryland", xlab="year", ylab='pm2.5 amount')
    axis(1, at=1:4, labels=names(emission_summary))
dev.off()

```

### Question 3
Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question.

http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/#with-x-axis-treated-as-categorical

```{r Question 3}

emission_summary <- sapply(split(baltimore,list(baltimore$year, baltimore$type)), function(grp) sum(grp$Emission, na.rm=TRUE))

library(reshape2)
temp <- melt(baltimore, id.vars=c('year', 'type'), measure.vars='Emissions')
emission_summary <- dcast(temp, year+type ~ variable, sum)

library(ggplot2)

png("Plot3.png")
ggplot(data= emission_summary, aes(factor(year), Emissions, group=type, color=type)) +
    geom_point(size=3) +                          # adding points
    geom_line(size=0.75, aes(linetype=type)) +    #adding line, and linetype for type 
    ggtitle("Annual pm2.5 Emission \n in Baltimore City, Maryland") +
    xlab("YEAR") + ylab("pm2.5 Emission amount")

dev.off()
```



### Question 4
Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?

```{r Question 4}

# find the SCC code related to coal
a <- grep("*coal*", SCC$Short.Name, ignore.case=TRUE)
scc_coal <- SCC[a,"SCC"]

# subset the NEI data with emissions from coal combustion related sources
coal_emission <- subset(NEI, SCC %in% scc_coal)


# fips is the location

library(reshape2)
temp <- melt(coal_emission, id.vars=c('year'), measure.vars='Emissions')
emission_summary <- dcast(temp, year ~ variable, sum)

#insert 1999 and 2008 values


library (ggplot2)

png("Plot4.png")
ggplot(data= emission_summary, aes(factor(year), Emissions, fill=factor(year))) +
    geom_bar(stat="identity") +        # to make Emissions as categorical
    ggtitle("pm2.5 Emission \nfrom coal combustion related sources \nacross the United States \n ") +
    xlab("Year") + ylab("pm2.5 Emission amount") +
    theme(legend.position = "none")
dev.off()

#    theme(legend.position = "bottom") +
#    scale_fill_discrete(name="")
       
```

### Question 5
How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?


```{r Question 5}

# find the SCC code for emissions from motor vehicle
a <- grep("*Vehicles*", SCC$EI.Sector, ignore.case=TRUE)
scc_vehicle <- SCC[a,"SCC"]

# extract observations of emission from motor vehicle sources in Baltimore City
baltimore_vehicle <- subset(NEI,fips=="24510" & SCC %in% scc_vehicle)

# merge SCC$SCC.Level.Three for motor vehicle sources
baltimore_vehicle <- merge(baltimore_vehicle, SCC[,c("SCC","SCC.Level.Three")], by="SCC")

# summarize the emission amount by year + vehicle
library(reshape2)
temp <- melt(baltimore_vehicle, id.vars=c('year','SCC.Level.Three'), measure.vars='Emissions')
emission_summary <- dcast(temp, year + SCC.Level.Three ~ variable, sum)


###### need to calculate the difference year over year

library (ggplot2)

veh_type <- factor(emission_summary$SCC.Level.Three)

png("Plot5.png")
ggplot(data= emission_summary, aes(factor(year), Emissions, group=SCC.Level.Three, color=veh_type)) +
    geom_point() +                          # adding points
    geom_line() +    #adding line
    ggtitle("pm2.5 Emission \nfrom motor vehicle sources \nin Baltimore \n ") +
    xlab("YEAR") + ylab("pm2.5 Emission amount") +
    scale_color_discrete(name= "motor vehicle type") +     theme(legend.text = element_text(size=6), legend.position = "bottom")

dev.off()


ggplot(data= emission_summary, aes(factor(year), Emissions, fill=factor(SCC.Level.Three))) +
    geom_bar(stat="identity") +        # to make Emissions as categorical
    ggtitle("pm2.5 Emission \nfrom motor vehicle sources \nin Baltimore") +
    xlab("Year") + ylab("pm2.5 Emission amount") +
    scale_fill_discrete(name= "motor vehicle type")
```

### Question 6

Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?

```{r Question 6}

# find the SCC code for emissions from motor vehicle
a <- grep("*Vehicles*", SCC$EI.Sector, ignore.case=TRUE)
scc_vehicle <- SCC[a,"SCC"]


# extract observations of emission from motor vehicle sources in Baltimore City (24510) and Los Angeles County (06037)
emission_vehicle <- subset(NEI,fips %in% c("24510","06037") & SCC %in% scc_vehicle)

# summarize the emission amount by location + year
library(reshape2)
temp <- melt(emission_vehicle, id.vars=c('fips','year'), measure.vars='Emissions')
emission_summary <- dcast(temp, fips+year ~ variable, sum)

png("Plot6.png")
ggplot(data= emission_summary, aes(factor(year), Emissions, fill=fips)) +
    geom_bar(stat="identity", position=position_dodge()) +        # to make Emissions as categorical
    ggtitle("pm2.5 Emission \nfrom motor vehicle sources") +
    xlab("Year") + ylab("pm2.5 Emission amount") +
    theme_bw() +
    scale_fill_hue(name= "Location", labels=c("Los Angeles County", "Baltimore, Maryland"))
dev.off()

```

