---
title: "Week4Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Download the file ProgAssignment3-data.zip file containing the data for Programming Assignment 3 from
the Coursera web site. Unzip the file in a directory that will serve as your working directory. 

The data for this assignment come from the Hospital Compare web site (http://hospitalcompare.hhs.gov)
run by the U.S. Department of Health and Human Services. The purpose of the web site is to provide data and
information about the quality of care at over 4,000 Medicare-certified hospitals in the U.S. This dataset essentially
covers all major U.S. hospitals. This dataset is used for a variety of purposes, including determining
whether hospitals should be fined for not providing high quality care to patients (see http://goo.gl/jAXFX
for some background on this particular topic).

The Hospital Compare web site contains a lot of data and we will only look at a small subset for this
assignment. The zip file for this assignment contains three files
• outcome-of-care-measures.csv: Contains information about 30-day mortality and readmission rates
for heart attacks, heart failure, and pneumonia for over 4,000 hospitals.
• hospital-data.csv: Contains information about each hospital.
• Hospital_Revised_Flatfiles.pdf: Descriptions of the variables in each file (i.e the code book).

A description of the variables in each of the files is in the included PDF file named Hospital_Revised_Flatfiles.pdf.

This document contains information about many other files that are not included with this programming
assignment. You will want to focus on the variables for Number 19 (“Outcome of Care Measures.csv”) and
Number 11 (“Hospital Data.csv”). You may find it useful to print out this document (at least the pages for
Tables 19 and 11) to have next to you while you work on this assignment. In particular, the numbers of
the variables for each table indicate column indices in each table (i.e. “Hospital Name” is column 2 in the
outcome-of-care-measures.csv file).

```{r cars}
#Step 0. downolaod the zip file and read the raw data
temp <- tempfile()
temp2 <-tempfile()


download.file("https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2FProgAssignment3-data.zip", temp)
unzip(temp,list=TRUE)
unzip(zipfile=temp, exdir=temp2)

```

## 1 Plot the 3--day mortality rates for heart attach

```{r}
outcome <- read.csv(file.path(temp2,'outcome-of-care-measures.csv',fsep='\\'), colClasses="character")
head(outcome)
ncol(outcome)
names(outcome)

# to make a simple histogram of the 30-day death rates from heart attack (column 11 in the outcome dataset)
temp_data <- as.numeric(outcome[,11])
hist(temp_data)

```

## 2 Finding the best hospital in a state
Write a function called **Best** that take 2 arguments: 
* the 2 character abbreviated name of a state
* an outcome name

The function reads the outcome-of-care-measures.csv file and returns a character vector with the name of the hospital that has the vest (i.e. lowest) 30-day mortality for the specified outcome in that state.
The hospital name is the name provided in the Hospital.Name variable.
The oucomes can be one of the:
* heart attack
* heart failure
* pneumonia
Hospitals that do not have data on a particular outcome should be excluded from the set of hospitals when deciding the rankings.

**Handling ties** If there is a tie for the best hospital for a given outcome, then the hospital names hsould be sorted in alphabetical order and the first hospital in that set should be chosen.
```{r}
best <- function(in_state, outcome) {
    
    ## Read outcome data
    read_data <- read.csv(file.path(temp2,'outcome-of-care-measures.csv',fsep='\\'), colClasses="character")
    
    ## check state and outcome are valid
    if (!(in_state %in% read_data$State)) {
        return ('invalid state')
    }
    
    compare = c('heart attack', 'heart failure', 'pneumonia')
    if (!(outcome %in% compare)) {
        return ('invalid outcome')
    }
    
    ## return hospital name in that sate with lowest 30-day death rate
    ### [2] "Hospital.Name"
    ### [7] "State"
    ### [11]"Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack"    
    ### [17] "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure" 
    ### [23] "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia"  
    
    a <- read_data[, c(2,7,11,17,23)]
    suppressWarnings(as.numeric)
    a[,c(3:5)] <- lapply(a[,c(3:5)],suppressWarnings(as.numeric))
    names(a) <- c('hospital','state','heart attack','heart failure','pneumonia')
    
    
    state_group <- split(a, a$state)
    b <-lapply(state_group, function (state_group) which(state_group[outcome]== min(state_group[outcome], na.rm=TRUE)))
    outcome_a = c('hospital',"state", outcome)
    as.data.frame(state_group[[in_state]])[b[in_state][[1]],outcome_a]
    
    ## double [[]] to get the element inside a list
}
   
```

## 3 Ranking hospitals by outcome in state
Write a function called rankhospital that takes 3 arguments:
* 2 character abbreviated name of a state
* an outcome
* ranking of a hospital in that state for that outcome.

the function will return a character vector containin the name of the hospital with the n-th outcome.
the num argument can take values 'best', 'worst' or integer indicate the ranking.
If the given number is larger than the number of hospitals in the state, the function returns NA.

When there is a tie, the tiem should be broken by using the hospital name ascendening order

```{r}

rankhospital <- function(state, outcome, num="best") {
    ## Read outcome data
    read_data <- read.csv(file.path(temp2,'outcome-of-care-measures.csv',fsep='\\'), colClasses="character")
    
    ## check state and outcome are valid
    if (!(state %in% read_data$State)) {
        return ('invalid state')
    }
    
    compare = c('heart attack', 'heart failure', 'pneumonia')
    if (!(outcome %in% compare)) {
        return ('invalid outcome')
    }
    
    ## return hospital name in that sate with lowest 30-day death rate
    ### [2] "Hospital.Name"
    ### [7] "State"
    ### [11]"Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack"    
    ### [17] "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure" 
    ### [23] "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia"  
    
    a <- read_data[, c(2,7,11,17,23)]
    a[,c(3:5)] <- lapply(a[,c(3:5)],suppressWarnings(as.numeric))
    names(a) <- c('hospital','state','heart attack','heart failure','pneumonia')
 
    state_group <- split(a, a$state)
    state_data <- state_group[[state]]
    
    state_data <- state_data[order(state_data[[outcome]], state_data$hospital, na.last = TRUE),]
    
    if (num =='best') {
        rank = 1
    } else if (num == 'worse') {
        rank = nrow(state_group[[state]]) - sum(is.na(state_data[[outcome]]))
    } else {
        rank = num
    }
    print (state_data[rank,'hospital'])
}

```

## Ranking hospital in all states
Write a function called rankall that take 2 arguments
* an outcome name (outcome)
* a hospital ranking

The function reads the outcome-of-care-mearsures.csv file
return a 2-column data frame containing the hospital in each state taht has the ranking specified in num.
For example the function call **rankall('heart attack','best')** would return a data frame containing the names of the hospitals that are the best in their repsective state. 
Hospital that do not have data on a particular outcome should be excluded from the set of hospitals when deciding the rankings.

```{r}

rankall <- function (outcome, num="best") {
    
    ## Read outcome data
    read_data <- read.csv(file.path(temp2,'outcome-of-care-measures.csv',fsep='\\'), colClasses="character")
    
    
    compare = c('heart attack', 'heart failure', 'pneumonia')
    if (!(outcome %in% compare)) {
        return ('invalid outcome')
    }
    
    ## for each state, find the hospital of the given rank
    
    a <- read_data[, c(2,7,11,17,23)]
    a[,c(3:5)] <- lapply(a[,c(3:5)],suppressWarnings(as.numeric))
    names(a) <- c('hospital','state','heart attack','heart failure','pneumonia')
 
    state_group <- split(a, a$state)
    
    options(stringsAsFactors = FALSE)
    result <- data.frame(hospital=character(),state=character(), stringsAsFactors = FALSE)

    for (i in seq_along(state_group)) {
        state_group[[i]] <- state_group[[i]][order(state_group[[i]][[outcome]], state_group[[i]]$hospital, na.last=TRUE),]
        
        if (num =='best') {
            rank = 1
        } else if (num == 'worse') {
            rank = nrow(state_group[[i]]) - sum(is.na(state_group[[i]][[outcome]]))
        } else {
            rank = num
        }
        
        #print(paste(state_group[[i]]$state[[1]], rank, sep=" :  "))
        result <- rbind(result, list(hospital = state_group[[i]][rank,'hospital'],
                                      state = names(state_group[i]))
                        )
    }
    return(result)
}

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
