---
title: "Statistical Inference Project"
author: "April L"
date: "April 7, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

#Overview  
In this project, you will investigate the exponential distribution in R
and compare it with the Central Limited Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter.

The mean of exponential distribution is 1/lambda,
and the standard deviation is also 1/lambda.
Set lambda = 0.2 for all the simulations.


##Simulation:
We simulate 1000 'average of 40 random exponentials observations'


```{r, echo=FALSE}

n <-1000
lambda <- 0.2
obs <- 40
means<- numeric(n)
variances <- numeric(n)

for (i in seq(1,n)) {
    sample <- rexp(obs,lambda)
    means[i] <- mean(sample)
    variances[i] <-var(sample)
}
 
```

### Sample Mean vs Theoretial Mean = 1/lambda
1. Show the sample mean and compare it to the theoretial mean of the distribution.

```{r}

sample_mean <- mean(means)
print(paste('The sample mean = ', round(sample_mean,2), ' is comparable to the theoretial mean of the distbution ' , 1/lambda, sep=''))

```


### Sample Variance vs. Theoretial Variance = (1/lambda)^2
2. Show how variable the sample (via variance) and compare it to the theoretiacal variance of the distribution.

```{r}
sample_var <- mean(variances)
print(paste('The sample mean = ', round(sample_var,2), ' is comparable to the theoretial mean of the distbution ' , (1/lambda)^2, sep=''))

```


### Distribution  
3. Show the distribution is approximately normal.

In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.

```{r}
par(mfrow=c(1,2))

# a large collection of random exponentials of lambda = 0.2
Norm <- rnorm(1000, mean=1/lambda, sd = sqrt(1/lambda))
hist(Norm, breaks=100, main='Distribution of 1000 random exponentials', col='red')

# distribution of a large collection of averages of 40 exponentials
hist(means, breaks=100, main="Distribution of 1000 averages of 40 exponentials, lambda = 0.2", col='blue')

```
Compare the 2 charts, it shows that the distribution of the means is approximately normal.

