---
title: "Statistic Inference Project Part 2"
date: "April 7, 2018"
output:
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(gridExtra)
```
###Part 2: Basic Inferential Data Analysis Instructions
Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.

####1. Load the ToothGrowth data 
```{r}
data("ToothGrowth")
data <- ToothGrowth
```
####2. Provide a basic summary of the data
```{r}
a<-ggplot( aes(x=supp, y=len), data=data) + geom_point(aes(color=supp))
b <- ggplot( aes(x=factor(dose), y=len), data=data) + geom_boxplot(aes(fill=factor(dose)))

grid.arrange(a, b, ncol=2, heights = 2)
summary(data)
``` 

###3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. ####3.1 Compare tooth growth by supp

H0: Toothgrowth is the same from VC supp and OJ supp (the 2 delivery methods: ascorbic acid and orange juice)  
H1: Toothgrowth is different 
```{r}
t.test(data=data, len~supp, paired=FALSE, var.equal=TRUE)

```

The P value = 0.06, which is > the signifance level, 0.05  
In addition, hypothesis value 0 is inside the 95% confidence interval.  
Thus,we do not reject the null hypothesis.
####3b. Compare the toothgrowth by dose
H0: Toothgrowth varies by dose  
HA: The different in dose has no impact to Toothgrowth 
```{r}
dose05 <- data[data$dose==0.5,]
dose1 <- data[data$dose==1,]
dose2 <- data[data$dose==2,]

t.test(dose05$len,dose1$len, paired=FALSE, alternative = "two.sided")
t.test(dose1$len,dose2$len, paired=FALSE, alternative = "two.sided")
t.test(dose05$len,dose2$len, paired=FALSE, alternative = "two.sided")

```
Compare the dose samples, all p-values are close to 0.  
None of the confidence interval has 0 in it.  
Thus we reject H0.  

###4. Conclusion:
Different ways to supply the vitamin C has no impact to tooth growth.  
It is possible that the difference in dose has an impact to tooth group.

